---
title: "Spatial Statistics - Introduction"
author: "Alexandra Pasternakova"
date: "11/12/2020 updated `r format(Sys.time(), '%B %d, %Y')`" 
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1: Simple spatial principles

```{r point generation, eval = TRUE}
n = 200

xmin <- 0
xmax <- 1
ymin <- 0
ymax <- 2

x = runif(n, xmin, xmax)
y = runif(n, ymin, ymax)

```


# Task 2: Plotting areas

```{r rectangle, eval=TRUE}
plot(x,y)

ls.str()

mapxy <- function(a = NA){
  plot(x, y, asp = a)
  rect(xmin, ymin, xmax, ymax)
}

mapxy(1:1)

```

# Task 3: Uniform pattern in a circle

```{r circle, results = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
# Load the spatstat package
library(spatstat)

```

```{r}

# Create this many points, in a circle of this radius
n_points <- 300
radius <- 10

# Generate uniform random numbers up to radius-squared using the objects above
r_squared <- runif(n_points, 0, radius^2)
angle <- runif(n_points, 0, 2*pi)

# Take the square root of the values to get a uniform spatial distribution
x <- sqrt(r_squared) * cos(angle)
y <- sqrt(r_squared) * sin(angle)

plot(disc(radius)); points(x, y)

```

# Task 4: Quadrat count test for randomness 

* **Does the p-value let you reject the null hypothesis?**

```{r quadrat, eval = TRUE, warning = FALSE}
# Some variables have been pre-defined
ls.str()
x <-  runif(300, -10,10)
y <-  runif(300, -10,10)
radius <- 10

summary(x)
# Set coordinates and window
ppxy <- ppp(x = x, y = y, window = disc(radius))

# Test the point pattern
qt <- quadrat.test(ppxy)

# Inspect the results
plot(qt)
print(qt)

```

**p-value of above 0.05 means we cannot reject the null hypothesis (null hypothesis: points are not spatially random)**

# Task 5: Creating a random point pattern with spatstat

```{r ppoisson, eval=TRUE}
# Create a disc of radius 10
disc10 <- disc(radius)

# Compute the rate as count divided by area
lambda <- 500 / area.owin(disc10)

# Create a point pattern object
ppois <- rpoispp(lambda = lambda, win = disc10)

# Plot the Poisson point pattern
plot(ppois)

npoints(ppois) #number of points generated
```

<br>

# Task 6: Simulating clustered and inhibitory patterns

```{r clustering,eval= TRUE, warning = FALSE}
# Create a disc of radius 10
disc10 <- disc(radius)

# Generate clustered points from a Thomas process
set.seed(123)
p_cluster <- rThomas(kappa = 0.35, scale = 1, mu = 3, win = disc10)
plot(p_cluster)

# Run a quadrat test
quadrat.test(p_cluster, alternative = "clustered") #p_value agrees it is clustered

# Regular points from a Strauss process
set.seed(123)
p_regular <- rStrauss(beta = 2.9, gamma = 0.025, R = .5, W = disc10)
plot(p_regular)

# Run a quadrat test
quadrat.test(p_regular, alternative = "regular") #p_value agrees it is regular, not clustered
```

<br>
<br>
<br>

# Task 7: Point Pattern testing

Two `ppp` objects have been created called `p1` and `p2` in your /data folder. Test if they are clustered or regular, and select the right answer. You'll need to run `quadrat.test` four times, once for each process with each hypothesis.

```{r, include=FALSE, eval=TRUE}
p1 <- readRDS("C:\\Users\\alexp\\Documents\\School\\Spatial_analytics\\HW\\HW3\\cds-spatial\\data\\p1.rds")
p2 <- readRDS("C:\\Users\\alexp\\Documents\\School\\Spatial_analytics\\HW\\HW3\\cds-spatial\\data\\p2.rds")
```

```{r, eval=TRUE, warning = FALSE}
quadrat.test(p1, alternative = "clustered") #p1 is clustered
```

```{r, eval=TRUE, warning = FALSE}
quadrat.test(p1, alternative = "regular") # confirmed p1 is not regular
```

```{r, eval=TRUE, warning = FALSE}
quadrat.test(p2, alternative = "clustered") #p2 is clustered
```

```{r, eval=TRUE, warning = FALSE}
quadrat.test(p2, alternative = "regular") #confirmed p2 is not regular
```

### Possible Answers

1. p1 is regular, p2 is regular
2. p1 is regular, p2 is clustered 
3. p1 is clustered, p2 is clustered
4. p1 is clustered, p2 is regular

**asnwer 3 is correct**
