---
title: "Spatial statistics: Point Patterns exercises"
author: "Alexandra Pasternakova"
date: "11/12/2020 updated `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Task 1: Nearest-neighbor distributions


```{r create points, echo = TRUE, warning = FALSE, message=FALSE}
library(spatstat)
```
```{r, eval = TRUE, echo = TRUE, warning = FALSE, results='hide'}
disc10 <- disc(radius = 10)
# p_poisson 
# Compute the rate as count divided by area
lambda <- 300/area.owin(disc10)
# Create a point pattern object
p_poisson <- rpoispp(lambda = lambda, win = disc10)
# p_cluster and p_regular
# Generate clustered points from a Thomas process
set.seed(123)
p_cluster <- rThomas(kappa = 0.35, scale = 1, mu = 3, win = disc10)
# Regular points from a Strauss process
set.seed(123)
p_regular <- rStrauss(beta = 2.9, gamma = 0.025, R = .5, W = disc10)


#plots
attach(mtcars)
par(mfrow=c(1,3))
plot(p_poisson)
plot(p_cluster)
plot(p_regular)
```


```{r NNA exercise, eval= TRUE, warning=FALSE}
# Point patterns are pre-defined
p_poisson; p_regular

# Calc nearest-neighbor distances for Poisson point data
nnd_poisson <- nndist(p_poisson)

# Draw a histogram of nearest-neighbor distances
hist(nnd_poisson)

# Estimate G(r)
G_poisson <- Gest(p_poisson)

# Plot G(r) vs. r
plot(G_poisson)

# Repeat for regular point data
nnd_regular <- nndist(p_regular)
hist(nnd_regular)
G_regular <- Gest(p_regular)
plot(G_regular)
```


# Task 2: Other point pattern distribution functions - multiscalar


```{r K function, eval= TRUE, warning=FALSE}
# Point patterns are pre-defined
p_poisson; p_cluster; p_regular

# Estimate the K-function for the Poisson points
K_poisson <- Kest(p_poisson, correction = "border")

# The default plot shows quadratic growth
plot(K_poisson, . ~ r)

# Subtract pi * r ^ 2 from the Y-axis and plot
plot(K_poisson, . - pi * r ^ 2 ~ r)

# Compute envelopes of K under random locations
K_cluster_env <- envelope(p_cluster, Kest, correction = "border")

# Insert the full formula to plot K minus pi * r^2
plot(K_cluster_env, . - pi * r ^ 2 ~ r)

# Repeat for regular data
K_regular_env <- envelope(p_regular, Kest, correction = "border")

plot(K_regular_env, . - pi * r ^ 2 ~ r)
```


# Task 3: Tree location pattern
Load a `ppp` object called `redoak` which contains the locations of trees in a woodland. Plot the tree locations and then test the hypothesis that the points are clustered against a null hypothesis of a random uniform distribution.

```{r, eval= TRUE, echo = FALSE}
redoak <- readRDS("C:\\Users\\alexp\\Documents\\School\\Spatial_analytics\\HW\\HW3\\cds-spatial\\data\\redoak.rds")
```

## Possible Answers

* I cannot reject the null hypothesis. 
* I can reject the null hypothesis with p-value less than 0.1 
* I can reject the null hypothesis with p-value less than 0.01

### Hint: check out the measure from the first exercise 
```{r Tree location pattern, eval= TRUE, warning=FALSE}
plot(redoak)
quadrat.test(redoak, alternative = "clustered")
```

**I can reject the null hypothesis with p-value less than 0.01** is correct

# 3) Optional: Using the mounds dataset from the Kazanlak Valley in data folder, what is pattern in these mounds over a range of distances (use K function)?

```{r Load data, eval = TRUE, echo = FALSE, warning = FALSE, message=FALSE, results='hide'}
library(sf)
library(maptools)
mounds <- st_read("C:/Users/alexp/Documents/School/Spatial_analytics/HW/HW3/cds-spatial/data/KAZ_mounds.shp")
```

```{r Mounds, eval= TRUE, warning=FALSE}
plot(st_geometry(mounds))

# convert to (point pattern) object - the study window seems to be in the dataframe already so I do not have to specify it
pp1 <- as.ppp( mounds)
plot(pp1)

# Estimate the K-function for the mound points
K_mounds <- Kest(pp1, correction = "border")

# default plot
plot(K_mounds, . ~ r)

# Subtract pi * r ^ 2 from the Y-axis and plot
plot(K_mounds, . - pi * r ^ 2 ~ r)

# Compute envelopes of K under random locations
K_Mounds_env <- envelope(pp1, Kest, correction = "border")

# Insert the full formula to plot K minus pi * r^2
plot(K_Mounds_env, . - pi * r ^ 2 ~ r)

```

**K values greater than K-expected indicate clustering of points at a given distance band**

```{r}
quadrat.test(pp1, alternative = "clustered")
```

**p-value < 2.2e-16 would confirm our findings, that the mounds are very clustered!!**