---
title: "HW4"
output: pdf_document
---

```{r, include=FALSE} 
options(tinytex.verbose = TRUE)

knitr::opts_chunk$set(echo = TRUE)
```

GitHub link: https://github.com/Digital-Methods-HASS/AU580227_Pasternakova_Alexandra.git

1) Use R to figure out how many elements in the vector below are greater than 2.

rooms <- c(1, 2, 1, 3, 1, NA, 3, 1, 3, 2, 1, NA, 1, 8, 3, 1, 4, NA, 1, 3, 1, 2, 1, 7, 1, NA)

```{r}

rooms <- c(1, 2, 1, 3, 1, NA, 3, 1, 3, 2, 1, NA, 1, 8, 3, 1, 4, NA, 1, 3, 1, 2, 1, 7, 1, NA)
#creates vector

rooms_filtered <- rooms[!is.na(rooms)]
#creates new vector taking values of "rooms" without NA
(rooms_filtered)
#prints elements of the vector on the console

rooms_filtered <- rooms_filtered[rooms_filtered > 2]
#filters out all values greater than 2
(rooms_filtered)
#prints values

length(rooms_filtered)
#counts how many elements there is
#answer is 8


```

2) What type of data is in the 'rooms' vector?

```{r}

class(rooms)
#tells us the tupe of data (numeric, logical, string...)
#the data is numeric

str(rooms)
#another option which also includes length and sample of data

```

3) What is the result of running the median() function on the above 'rooms' vector?


```{r}

median(rooms)
#the result is unknown due to our data including NA vaues
#we could use the vector without NA values to count the median

median(rooms , na.rm = TRUE)
#by including na.rm we tell the program to remove NA values before starting computation
#result for median is 1.5


```
